A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-19 tidyindex.qmd
tidied up section 2

<!-- When indexes have a spatial and/or temporal component, different choices on how space and time is aggregated will introduce uncertainty. A slight change in the variable value can sometimes introduce a big difference for the index.  -->
# Background to index usage and development

@@

The OECD handbook [@oecd_handbook_2008] provides a framework of steps and recommendations. for composing socio-economic composite indexes. The drought index handbook [@svoboda2016handbook], instead, provides a practical guide outlining the advantage and disadvantage of various drought indexes, as recommended by World Meteorology Organization. Review papers in the literature summarise different methodologies for constructing indexes from different variables and statistical methods, for example, @zargar2011review, @hao_drought_2015 @alahacoon_comprehensive_2022 for drought indexes. In specific domains, attention are given to analyze the index construction process, including the implications of the choices made during the index construction [@jones_vulnerability_2007] and uncertainty or sensitivity analysis [@uncertainty; @tate_social_2012].

In software development, many tools encapsulate all the construction steps into a single function to allow computing indexes with provided data. This includes the `SPEI` package [@spei] for two drought indexes, the `gpindex` [@gpindex] package for price indexes, and the `fundiversity` package [@fundiversity] for functional diversity indexes in ecology. The package `COINr` [@COINr] deviates from this one-size-fit-all approach by following the guidelines proposed in the OECD handbook to construct, analyze, and visualize composite indexes.

>>>>

<!-- When indexes have a spatial and/or temporal component, different choices on how space and time is aggregated will introduce uncertainty. A slight change in the variable value can sometimes introduce a big difference for the index.  -->

# Background to index development

@@

There are many documents providing advice on how to construct indexes for different fields, and review articles describing the range of available indexes for specific purposes. The OECD handbook [@oecd_handbook_2008] provides a comprehensive guide for computing socio-economic composite indexes, with detailed steps and recommendations. The drought index handbook [@svoboda2016handbook] provides details of various drought indexes and recommendations from the World Meteorology Organization.  @zargar2011review, @hao_drought_2015 and @alahacoon_comprehensive_2022 are review papers describing the range of possible drought indexes.

There is also some attention being given to the diagnosis of indexes, and incorporation of uncertainty. @jones_vulnerability_2007 investigates the methodological choices made in the development of indexes for assessing vulnerable neighborhoods. @uncertainty describes incorporating uncertainty estimates and conducting sensitivity analysis on composite indexes. @tate_social_2012, similarly, makes a comparative assessment of social vulnerability indexes based on uncertainty estimation and sensitivity analysis. (XXX Something about Ursula's colleagues paper here too.)

There are also R packages supporting index calculation. The  `SPEI` package  [@spei] computes two drought indexes. The `gpindex` package [@gpindex] computes price indexes, and the `fundiversity` package [@fundiversity] computes functional diversity indexes for ecological study. The package `COINr` [@COINr] is more ambitious, making a start on following the broader guidelines in the OECD handbook to construct, analyze, and visualize composite indexes.

From reviewing this literature, and in the process of developing methods for making it easier to work with multivariate spatiotemporal data, it seems possible to think about indexes in a more organised, cohesive and standard manner. Actually, it seems that the area could benefit from a *tidy* approach.

### 2023-08-19 tidyindex.tex
tidied up section 2

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their value and variation in values. Here we discuss a
modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
with different the index with different parameter choices, adjust steps
in the index definition by adding, removing, and swapping them to
experiment with various index designs, calculate uncertainty measures,
and assess an index's robustness. The paper presents three examples to
illustrate the pipeline framework usage: comparison of two different
indexes designed to monitor the spatio-temporal distribution of drought
in Queensland, Australia; the effect of dimension reduction choices on
the Global Gender Gap Index (GGGI) on a country's ranking; and how to
calculate bootstrap confidence intervals for XXX index. The methods are
supported by a new R package, called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, interior hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{background-to-index-usage-and-development}{%
\section{Background to index usage and
development}\label{background-to-index-usage-and-development}}

The OECD handbook (OECD, European Union, and Joint Research Centre -
European Commission 2008) provides a framework of steps and
recommendations. for composing socio-economic composite indexes. The
drought index handbook (Svoboda, Fuchs, et al. 2016), instead, provides
a practical guide outlining the advantage and disadvantage of various
drought indexes, as recommended by World Meteorology Organization.
Review papers in the literature summarise different methodologies for
constructing indexes from different variables and statistical methods,
for example, Zargar et al. (2011), Hao and Singh (2015) Alahacoon and
Edirisinghe (2022) for drought indexes. In specific domains, attention
are given to analyze the index construction process, including the
implications of the choices made during the index construction (Jones
and Andrey 2007) and uncertainty or sensitivity analysis (Saisana,
Saltelli, and Tarantola 2005; Tate 2012).

In software development, many tools encapsulate all the construction
steps into a single function to allow computing indexes with provided
data. This includes the \texttt{SPEI} package (Vicente-Serrano,
Beguería, and López-Moreno 2010) for two drought indexes, the
\texttt{gpindex} (Martin 2023) package for price indexes, and the
\texttt{fundiversity} package (Grenié and Gruson 2023) for functional
diversity indexes in ecology. The package \texttt{COINr} (Becker et al.
2022) deviates from this one-size-fit-all approach by following the
guidelines proposed in the OECD handbook to construct, analyze, and
visualize composite indexes.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their value and variation in values. Here we discuss a
modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
with different the index with different parameter choices, adjust steps
in the index definition by adding, removing, and swapping them to
experiment with various index designs, calculate uncertainty measures,
and assess an index's robustness. The paper presents three examples to
illustrate the pipeline framework usage: comparison of two different
indexes designed to monitor the spatio-temporal distribution of drought
in Queensland, Australia; the effect of dimension reduction choices on
the Global Gender Gap Index (GGGI) on a country's ranking; and how to
calculate bootstrap confidence intervals for XXX index. The methods are
supported by a new R package, called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{background-to-index-development}{%
\section{Background to index
development}\label{background-to-index-development}}

There are many documents providing advice on how to construct indexes
for different fields, and review articles describing the range of
available indexes for specific purposes. The OECD handbook (OECD,
European Union, and Joint Research Centre - European Commission 2008)
provides a comprehensive guide for computing socio-economic composite
indexes, with detailed steps and recommendations. The drought index
handbook (Svoboda, Fuchs, et al. 2016) provides details of various
drought indexes and recommendations from the World Meteorology
Organization. Zargar et al. (2011), Hao and Singh (2015) and Alahacoon
and Edirisinghe (2022) are review papers describing the range of
possible drought indexes.

There is also some attention being given to the diagnosis of indexes,
and incorporation of uncertainty. Jones and Andrey (2007) investigates
the methodological choices made in the development of indexes for
assessing vulnerable neighborhoods. Saisana, Saltelli, and Tarantola
(2005) describes incorporating uncertainty estimates and conducting
sensitivity analysis on composite indexes. Tate (2012), similarly, makes
a comparative assessment of social vulnerability indexes based on
uncertainty estimation and sensitivity analysis. (XXX Something about
Ursula's colleagues paper here too.)

There are also R packages supporting index calculation. The
\texttt{SPEI} package (Vicente-Serrano, Beguería, and López-Moreno 2010)
computes two drought indexes. The \texttt{gpindex} package (Martin 2023)
computes price indexes, and the \texttt{fundiversity} package (Grenié
and Gruson 2023) computes functional diversity indexes for ecological
study. The package \texttt{COINr} (Becker et al. 2022) is more
ambitious, making a start on following the broader guidelines in the
OECD handbook to construct, analyze, and visualize composite indexes.

From reviewing this literature, and in the process of developing methods
for making it easier to work with multivariate spatiotemporal data, it
seems possible to think about indexes in a more organised, cohesive and
standard manner. Actually, it seems that the area could benefit from a
\emph{tidy} approach.
